The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2014-02-16 paper/ms.tex
Final Version for Mark

A basic understanding of the quantitative behavior of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. Analytic studies of configurations including
some kind of bulk flow only include the case of a sphere with a Hubble
like expansion flow \citep{LoebRybicki}.

@@

The central result of this paper is summarized in Figure
\ref{fig:differentvelocities}. It shows the evident
impact of rotation on the morphology of the emergent \ly line.
The panel focus on the results for all of our symmetric models
 $\tau_{H}=(10^{6}, 10^{6}, 10^{5}$), showing that the
influence of rotation is present both when the photons are either
homogeneously or centrally initialized over the gas volume.

@@

In order to quantify the transition to a single peak we measure the
position for the peak maxima as a function of the rotational
velocity. Our results are presented in Figure
\ref{fig:maximumsvsvelocity} where it is clear that for central
sources there are always two peaks with fixed positions as the rotational
velocity changes. However, in the case of homogeneously emitted
sources the maxima position remain close to constant until some
velocity threshold the line becomes single peaked with $x_{\rm m}=0$
\kms. By inspection we find that in all cases the transition has
occurred when the rotational velocity is larger than half width .

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f6.eps}
  \includegraphics[width=0.40\textwidth]{f7.eps}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel. Escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel. Escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$, where $\tau_{A}$ is taking into account the albedo. The    analytic solution for the infinite
    slab is shown as a continuous    line. Different lines corresponds
    to different rotational velocities for the homogeneous models.     \label{fig:efvsv}} 
\end{figure*}

@@

In the right panel of Figure \ref{fig:efvsv} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{A}$ where $\tau_{A} = (1 - A)\tau_{a}$ and $A$ is the albedo
, an approximation that is valid only
in the limit $a\tau_{\rm   H}\gg 1$. The dashed lines in the right panel of Figure
\ref{fig:efvsv} show the results for the different rotational
velocities for the homogeneous models. First of all we note that the escape fraction does not
increase significantly from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, a fact
that is explained since there is a transition from an opaque to an extremely opaque medium
which affects how the \ly photons escape from the medium.

@@

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{Verhamme06,DijkstraKramer}. In this
paper we show that gas bulk rotation can also be considered as a
probable origin for that behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

\item The line width increases with rotational velocity. The line with
  approximately following the functional for  ${\rm FWHM}^2 = {\rm FWHM}_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where FWHM$_{0}$ indicates the line
  width for the static case. $\lambda$ is a constant determined from
  the radiative transfer results it is $\lambda_{\rm c}=3.9\pm$ and
  $\lambda_{\rm h}=1.9\pm$ for the central and homogeneous source
  distribution, respectively.

>>>>

A basic understanding of the quantitative behaviour of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. Analytic studies of configurations including
some kind of bulk flow only include the case of a sphere with a Hubble
like expansion flow \citep{LoebRybicki}.

@@

The central result of this paper is summarized in Figure
\ref{fig:differentvelocities}. It shows the evident
impact of rotation on the morphology of the emergent \ly line.
The panel focus on the results for all of our symmetric models
 $\tau_{H}=(10^{5}, 10^{6}, 10^{7}$), showing that the
influence of rotation is present both when the photons are either
homogeneously or centrally initialized over the gas volume.

@@

In order to quantify the transition to a single peak we measure the
position for the peak maxima as a function of the rotational
velocity. Our results are presented in Figure
\ref{fig:maximumsvsvelocity} where it is clear that for central
sources there are always two peaks with fixed positions as the rotational
velocity changes. However, in the case of homogeneously emitted
sources the maxima position remain close to constant until some
velocity threshold the line becomes single peaked with $x_{\rm m}=0$
\kms. By inspection we find that in all cases the transition has
occurred when the rotational velocity is larger than half width.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f6.eps}
  \includegraphics[width=0.40\textwidth]{f7.eps}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel. Escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel. Escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
    slab is shown as a continuous    line. Different lines corresponds
    to different rotational velocities for the homogeneous models.     \label{fig:efvsv}} 
\end{figure*}

@@

In the right panel of Figure \ref{fig:efvsv} we put these results in the context of
the analytic solution for the infinite slab \citep{Neufeld90}. In
Neufeld's set-up the analytic solution depends uniquely on the product
$(a\tau_{\rm   H})^{1/3}\tau_{A}$ where $\tau_{A} = (1 - A)\tau_{a}$ , an approximation 
that is valid only in the limit $a\tau_{\rm   H}\gg 1$. The dashed lines in
 the right panel of Figure \ref{fig:efvsv} show the results for the different rotational
velocities for the homogeneous models. First of all we note that the escape fraction does not
increase significantly from $\tau_{\rm H}=10^{5}$ to $\tau_{\rm H}=10^{6}$, a fact
that is explained since there is a transition from an opaque to an extremely opaque medium
which affects how the \ly photons escape from the medium.

@@

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{Verhamme06,DijkstraKramer}. In this
paper we show that gas bulk rotation can also be considered as a
probable origin for that behaviour, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

\item The line width increases with rotational velocity. The line with
  approximately following the functional for  ${\rm FWHM}^2 = {\rm FWHM}_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where FWHM$_{0}$ indicates the line
  width for the static case. $\lambda$ is a constant determined from
  the radiative transfer results it is $\lambda_{\rm c}=3.9\pm0.041$ and
  $\lambda_{\rm h}=1.9\pm0.062$ for the central and homogeneous source
  distribution, respectively.
